Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-04 paper/paper.tex
new structure
full toggle, reorganization in progress

\usepackage{hyperref}
\usepackage{url}

@@

@@

There are several ways to build a programming language
on top of or besides another language.  The easiest is an
interpreter --- a program that reads a program, in its entirety
or line-by-line, and executes it by applying operational
semantics of a certain kind to the language. \textsc{Basic} is
famous for line-by-line interpreted implementations.

@@

On top of these two approaches are methods in which a new
language is implemented `inside' another language of the same
level of abstraction. Different languages provide different
means for this; Lisp is famous for the macro facility
that allows to extend the language almost without
restriction --- by writing \textit{macros}, one adds new
constructs to the existing language. There are several uses of
macros --- one is to extend the language \textit{syntax}, for
example, by adding new control structures; another is to keep
the existing syntax but alter the \textit{operational semantics}
--- the way programs are executed and compute their outputs.

@@

@@

@@

\subsection{Expression Kinds}
\label{sec:expr}

So far we introduced the basics of Anglican compilation to
Clojure. The described approaches and techniques are important
for grasping the language implementation but relatively well-known.
In the rest of the section we focus on challenges we met and
resolved while implementing Anglican, as well as on
implementation of unique features of Anglican as a probabilistic
programming language.

Continuations are functions that are called in tail positions
with the computed value and state as their arguments --- in CPS
there is always a function call in every tail position and never
a value. Continuations are passed to CPS transformers, and when
transformers are called recursively, the continuations are
generated on the fly.

\paragraph{Managing stack size}

In implementations of functional programming languages stack
growth is avoided through \textit{tail call optimization} (TCO).
However, Clojure does not support a general form of TCO, and
CPS-transformed code that creates deeply nested calls will
easily exhaust the stack. Anglican employs a workaround called
\textit{trampolining} --- instead of inserting a continuation
call directly, the transformer always wraps the call into a
\textit{thunk}, or parameterless function. The thunk is returned
and called by the trampoline (Clojure provides function
\texttt{trampoline} for this purpose) --- this way the
computation continues, but the stack is collapsed on every
continuation call. Function \texttt{continue} implements the
wrapping and is invoked on every continuation call:
\begin{lstlisting}[style=default]
=> (continue 'cont 'value 'state)
(fn [] (cont value state))
\end{lstlisting}
Correspondingly, the full, wrapped CPS form of
\begin{lstlisting}[style=default]
(fn [x y] (+ x y))
\end{lstlisting}
becomes
\begin{lstlisting}[style=default]
(fn [cont $state x y]
  (fn []
    (cont (+ x y) $state)))
\end{lstlisting}
When the CPS-transformed function is called, it returns a
\textit{thunk} (a parameterless function) which is then
re-invoked through the trampoline, with the stack collapsed.

\paragraph{Avoiding exponential code growth}

To circumvent this inefficiency, CPS transformers for
expressions with multiple continuation points (\texttt{if} and
derivatives, \texttt{and}, \texttt{or}, and \texttt{case}) bind
the continuation to a fresh symbol if it is not yet a symbol.
Macro \texttt{defn-with-{\linebreak[0]}named-{\linebreak[0]}cont}
establishes the binding automatically:
\begin{lstlisting}[style=default]
=> (cps-of-if '(c t f) '(fn [x] (* 2 x)))
(let [cont (fn [x] (* 2 x))]
  (if c
      (fn [] (cont t $state))
      (fn [] (cont f $state))))
\end{lstlisting}

When an Anglican function is transformed into a Clojure function
by \texttt{fn-cps}, two auxiliary parameters are added to the
beginning of the parameter list --- continuation and state.
Correspondingly, when a function \textit{call} is transformed
(by \texttt{cps-of-application} or \texttt{cps-of-apply}), the
current continuation and the state are passed to the called
function. Anglican can also call Clojure functions; however,
Clojure functions do not expect these auxiliary parameters.
To allow the mixing of Anglican (CPS-transformed) and Clojure
function calls in Anglican code, the Anglican compiler must be
able to recognize `primitive' (that is, implemented in Clojure
rather than in Anglican) functions.

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime. However, this would impact performance, and
good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is an Anglican function. Global dynamically-bound 
variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespaces*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure primitive function that takes the first
element from an ordered collection such as list and vector,
but inside the let block in the
following example, \texttt{first} is an Anglican function:
\begin{lstlisting}[style=default]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

% A common pattern when programming with random processes in 
% Anglican is to generate a sequences of sampled values and 
% constructing an updated process at each step

@@

The description of Anglican language and environment would be
incomplete without a case study showing the process of building
and executing a solution of an inference problem. This case
study takes a problem for which a solution is not immediately
obvious, \textit{the Deli dilemma}, and guides through
writing an Anglican program for the problem, executing the
program, and post-processing results.

@@

>>>>

\usepackage{hyperref}
\usepackage{url}
\usepackage{etoolbox}

@@

\newtoggle{full}
\togglefalse{full}

@@

\iftoggle{full}{%
There are several ways to build a programming language
on top of or besides another language.  The easiest is an
interpreter --- a program that reads a program, in its entirety
or line-by-line, and executes it by applying operational
semantics of a certain kind to the language. \textsc{Basic} is
famous for line-by-line interpreted implementations.

@@

On top of these two approaches are methods in which a new
language is implemented `inside' another language of the same
level of abstraction. 
}{%
One way to implement a programming language is to embed it within another language.
}%
Different languages provide different
means for this; Lisp is famous for the macro facility
that allows to extend the language almost without
restriction --- by writing \textit{macros}, one adds new
constructs to the existing language. There are several uses of
macros --- one is to extend the language \textit{syntax}, for
example, by adding new control structures; another is to keep
the existing syntax but alter the \textit{operational semantics}
--- the way programs are executed and compute their outputs.

@@

\section{Related Work}
\label{sec:related}

{\it TODO}

@@

\subsection{Probabilistic Forms}

{\it TODO}

@@

@@

Before delving into implementation details, we provide a case
study of using the Anglican language and environment. This case
study takes a problem for which a solution is not immediately
obvious, \textit{the Deli dilemma}, and guides through writing
an Anglican program for the problem, executing the program, and
post-processing results.

@@

\iftoggle{full}{%
\subsection{Expression Kinds}
\label{sec:expr}

}{%
}

\iftoggle{full}{%
So far we introduced the basics of Anglican compilation to
Clojure. The described approaches and techniques are important
for grasping the language implementation but relatively well-known.
}{%
Anglican compilation to Clojure is based on CPS transformation
and compilation techniques, many of which are
well known and widely used~\cite{AJ89,A07}.
}%
In the rest of the section we focus on challenges we met and
resolved while implementing Anglican, as well as on
implementation of unique features of Anglican as a probabilistic
programming language.

\iftoggle{full}{
Continuations are functions that are called in tail positions
with the computed value and state as their arguments --- in CPS
there is always a function call in every tail position and never
a value. Continuations are passed to CPS transformers, and when
transformers are called recursively, the continuations are
generated on the fly.

\paragraph{Managing stack size}
}{%
Continuations may lead to unbounded stack growth in
recursive code.
}%
In implementations of functional programming languages stack
growth is avoided through \textit{tail call optimization} (TCO).
However, Clojure does not support a general form of TCO, and
CPS-transformed code that creates deeply nested calls will
easily exhaust the stack. Anglican employs a workaround called
\textit{trampolining} --- instead of inserting a continuation
call directly, the transformer always wraps the call into a
\textit{thunk}, or parameterless function. The thunk is returned
and called by the trampoline (Clojure provides function
\texttt{trampoline} for this purpose) --- this way the
computation continues, but the stack is collapsed on every
continuation call. Function \texttt{continue} implements the
wrapping and is invoked on every continuation call:
\begin{lstlisting}[style=default]
=> (continue 'cont 'value 'state)
(fn [] (cont value state))
\end{lstlisting}
Correspondingly, the full, wrapped CPS form of
\begin{lstlisting}[style=default]
(fn [x y] (+ x y))
\end{lstlisting}
becomes
\begin{lstlisting}[style=default]
(fn [cont $state x y]
  (fn []
    (cont (+ x y) $state)))
\end{lstlisting}
When the CPS-transformed function is called, it returns a
\textit{thunk} (a parameterless function) which is then
re-invoked through the trampoline, with the stack collapsed.

\iftoggle{full}{%
\paragraph{Avoiding exponential code growth}

To circumvent this inefficiency, CPS transformers for
expressions with multiple continuation points (\texttt{if} and
derivatives, \texttt{and}, \texttt{or}, and \texttt{case}) bind
the continuation to a fresh symbol if it is not yet a symbol.
Macro \texttt{defn-with-{\linebreak[0]}named-{\linebreak[0]}cont}
establishes the binding automatically:
\begin{lstlisting}[style=default]
=> (cps-of-if '(c t f) '(fn [x] (* 2 x)))
(let [cont (fn [x] (* 2 x))]
  (if c
      (fn [] (cont t $state))
      (fn [] (cont f $state))))
\end{lstlisting}
}{
}

When an Anglican function is transformed into a Clojure
function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary parameters are added to the
beginning of the parameter list --- continuation and state.
Correspondingly, when a function \textit{call} is
transformed\iftoggle{full}{ (by \texttt{cps-of-application} or
\texttt{cps-of-apply})}{}, the
current continuation and the state are passed to the called
function. Anglican can also call Clojure functions; however,
Clojure functions do not expect these auxiliary parameters.
To allow the mixing of Anglican (CPS-transformed) and Clojure
function calls in Anglican code, the Anglican compiler must be
able to recognize `primitive' (that is, implemented in Clojure
rather than in Anglican) functions.

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime. However, this would impact performance, and
good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is an Anglican function. Global dynamically-bound 
variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespaces*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure primitive function that takes the first
element from an ordered collection such as list and vector,
but inside the let block in the
following example, \texttt{first} is an Anglican function:
\begin{lstlisting}[style=default]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical
scope\iftoggle{full}{ (\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading}{.}

% A common pattern when programming with random processes in 
% Anglican is to generate a sequences of sampled values and 
% construct an updated process at each step.

\section{Performance}

{\it TODO}
